Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife

### 2023-04-06 analysis/paper/paper.Rmd
ongoing
to do - figure what is going wrong in closerLookAtPatterns script.

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021]. These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository (add link). We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.

@@

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy include switchback, stream crossings, inaccurate harvest locations. Missing switchbacks and too many stream crossings also reduce our total amount of roads. stream crossing behaviour could improved with more detailed high resolution cost. harvest and landing data. 
- density of landing pts has a larger effect than choice of road projection algorithm. So maybe don't get too worked up about the differences between one algorithm and another. if more accurate projections are required, focus first on refining cost & harvest landing locations.
- Selected revelstoke because it is topographically complex, so a good challenge. However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. Compare to Hardy results in eastern canadian shield - better road density projections in that landscape suggests better performance of simple iterative least cost path algorthms in more moderate terrain. 
- Recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For this example landscape projections of disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). Density is harder to project partly because none of these algorithms can project variation in density within 100 m pixels. If projection of impacts on wildlife is a goal, it may be helpful to consider including more easily projectable road predictors in wildlife response models (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.
-->
```
<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->

>>>>

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021]. These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository (https://osf.io/8vmqh/). We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>.

@@

```{=html}
<!-- Discussion pts
- simple algorithms, practical to apply over large areas with minimal available information (unlike operational models)
- as expected, not great at projecting exact road locations. sources of discrepancy highlighted in this mountainous terrain include switchbacks, stream crossings, inaccurate harvest locations. we selected revelstoke because it is topographically complex, so a good challenge. Hardy result in eastern Canadian shield shows better overall accuracy projecting road density in less extreme terrain when the costs of water crossings are more carefully accounted for. Stream crossing behaviour in Revelstoke landscape could improved with more a detailed high resolution cost surface. But our results highlight that performance of these algorithms depends not only on accounting for cost but on complexity of the terrain. In particular, none of these simple algorithms will project switchbacks on steep slopes. A related fundamental limitation is that none of these algorithms can project tortuosity within raster cells, so in complex landscapes projections of road density will become less accurate as raster cell resolution increases. 
- Aspatial overall performance metrics can be difficult to interpret. To more deeply understand variation in projection accuracy we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch clarified that some discrepancies in our landscape are due to errors in our input data, rather than projection errors. Better harvest & existing road data would allow us to distinguish errors in projections from errors in the input data (cite lit in coops paper)
 - Outside cutblocks, we see very little variation in performance among our method variants - any of them are better than nothing. Within cutblocks that also have observed roads, the density of landing points has a much larger impact on performance than the choice of DCLP or MST algorthm, and the Hardy algorithm more accurately projects road density than either. Hardy algorithm is also less prone to putting roads on block edges. If accurate road density within blocks is primary goal, consider using Hardy algorithm. Thus far only implemented in Landis II and QGIS extension, but source code is open. If there is a need could integrate into roads R package. Or consider simpler adjustments to algorithms for selecting landing points such as placing landing points away from cutblock edges. 
- However, before investing in more complexity recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor, and at a landscape scale a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates (). In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are pretty good (and better than nothing). If our wildlife response models depend on distance to road or buffered disturbance footprint, these somewhat crude road projections are likely good enough.
- A related point is that it may helpful to recognize that some road metrics are easier to project than others when constructing model to project wildlife responses to disturbance ... (Bodner,). If interested in predictors that we did not consider, see example script. Can evaluate alternative metrics without rerunning the projections.

However, This example projected new roads over fairly short distances, new landings were always fairly close to existing roads or other landings so there was not a lot of space for the projection to go wrong. 
-->
```
<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->
